---
title: "Data format"
author: "Richard Aubrey White"
date: "2022-01-19"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r echo=FALSE, include=FALSE}
library(data.table)
library(magrittr)
```

## splfmt_rts_v1_mini / splfmt_rts_v1_max

This document presents the data format `splfmt_rts_v1_mini`.

`splfmt_rts_v1_mini` is the data format that the Sykdomspulsen team recommends using for the real-time surveillance of infectious diseases.

`splfmt_rts_v1_max` is the internal data format for the real-time analysis and disease surveillance framework Sykdomspulsen Core. `splfmt_rts_v1_max` is identical to `splfmt_rts_v1_mini`, except that it contains a few extra (superfluous) columns that make it easier to perform data selections on large datasets. It is easy to convert between `splfmt_rts_v1_mini` and `splfmt_rts_v1_max`.

## Style

**Language**

English is the primary language for our code.

Names that are abbreviations or in Norwegian are kept as they are: data sources such as `msis, dar, sysvak, normomo`.

**Capital letters**

Capital letters are to be avoided whenever possible. This is also the case in filenames (e.g. `data.rds` is preferred to `data.RDS`)

**snake_case or camelCase?**

Use snake_case.

**Timestamping of file names**

In results (e.g. reports), an indicator of time when the files are created are necessary. It allows us to find which one is the most recent version of files with the same names, and it allows easy tracking of an Airflow error. 

e.g. `Epidemiologisk_situasjonsrapport_2021-05-31_0659.docx` for a report generated on May 31, 2021 06:59 AM. 

## Ordering of variables

Sometimes variables need to be ordered. Variables should be ordered as follows:

- time
- location
- age
- sex

e.g. A database table could be called `msis_by_time_location_age_sex` or a filename could be called `2020_oslo_05-10_male.xlsx`

## Time

Time functions can be obtained from [spltime](/spltime). Missing time data should be coded as `NA`.

```{r echo=FALSE, results='asis'}
d <- rbind(
  data.frame(
    granularity_time = "day",
    class = "Date",
    fn = "as.Date",
    example = "2021-12-31"
  ),
  data.frame(
    granularity_time = "isoweek (numeric)",
    class = "numeric",
    fn = "spltime::isoweek_n",
    example = "1"
  ),
  data.frame(
    granularity_time = "isoweek (character)",
    class = "character",
    fn = "spltime::isoweek_c",
    example = "\"01\""
  ),
  data.frame(
    granularity_time = "isoyear (numeric)",
    class = "character",
    fn = "spltime::isoyear_n",
    example = "2021"
  ),
  data.frame(
    granularity_time = "isoyear (character)",
    class = "character",
    fn = "spltime::isoyear_c",
    example = "\"2021\""
  ),
  data.frame(
    granularity_time = "isoyearweek",
    class = "character",
    fn = "spltime::isoyearweek_c",
    example = "\"2021-01\""
  ),
  data.frame(
    granularity_time = "event_*_date1_to_date2",
    class = "character",
    fn = "as.character",
    example = "\"event_covid19_norway_vaccination_2020_12_02_to_9999_01_01\", \"event_covid19_norway_2020_02_21_to_9999_01_01\""
  )
)

gt::gt(d) %>%
  gt::tab_options(
    table.width = "1100px"
  ) %>%
  gt::tab_header(title = "Valid times in the splverse format") %>%
  gt::cols_label(
    granularity_time = "Time (Granularity)",
    class = "Class",
    fn = "Function",
    example = "Example(s)"
  ) %>%
  gt::cols_width(
    granularity_time ~ "20%",
    class ~ "15%",
    fn ~ "20%",
    example ~ "55%"
  ) %>%
  gt::tab_style(
    style = list(
      gt::cell_text(decorate = "line-through")
      ),
    locations = gt::cells_body(
      columns = gt::everything(),
      rows = stringr::str_detect(granularity_time, "character")
    )
  ) %>%
  gt::tab_footnote(
    footnote = "isoweek (numeric) is used when it is a standalone variable.",
    locations = gt::cells_body(
      columns = granularity_time,
      rows = granularity_time == "isoweek (numeric)"
    )
  ) %>%
    gt::tab_footnote(
    footnote = "isoweek (character) is only used within isoyearweek.",
    locations = gt::cells_body(
      columns = granularity_time,
      rows = granularity_time == "isoweek (character)"
    )
  ) %>%
  gt::tab_footnote(
    footnote = "isoyear (numeric) is used when it is a standalone variable.",
    locations = gt::cells_body(
      columns = granularity_time,
      rows = granularity_time == "isoyear (numeric)"
    )
  ) %>%
    gt::tab_footnote(
    footnote = "isoyear (character) is only used within isoyearweek.",
    locations = gt::cells_body(
      columns = granularity_time,
      rows = granularity_time == "isoyear (character)"
    )
  ) %>%
    gt::tab_footnote(
    footnote = "If the event is ongoing, then the 'to' date should be 9999_01_01.",
    locations = gt::cells_body(
      columns = granularity_time,
      rows = stringr::str_detect(granularity_time, "event")
    )
  )
```

## Location

Locations can be obtained from [spldata](/spldata). Valid locations (and location types) are available in `spldata::norway_locations_names()`.

```{r echo=FALSE, results='asis'}
d <- spldata::norway_locations_names()[,.(
  n=.N, 
  location_code = location_code[1],
  location_name = location_name[1],
  location_name_description_nb = location_name_description_nb[1],
  location_name_file_nb_utf = location_name_file_nb_utf[1],
  location_name_file_nb_ascii = location_name_file_nb_ascii[1]
  ),
  by=.(granularity_geo)
  ]

gt::gt(d) %>%
  gt::tab_options(
    table.width = "1500px"
  ) %>%
  gt::tab_header(title = "Valid locations and location types in the splverse format") %>%
  gt::cols_label(
    granularity_geo = "Geo (Granularity)",
    n = "N"
  ) %>%
  # gt::cols_width(
  #   granularity_time ~ "20%",
  #   class ~ "15%",
  #   fn ~ "20%",
  #   example ~ "55%"
  # ) %>%
  gt::tab_spanner(
    label = "Examples",
    columns = c(location_code, location_name, location_name_description_nb, location_name_file_nb_utf, location_name_file_nb_ascii)
  ) %>%
  gt::tab_footnote(
    footnote = gt::md("**location_code**: Used a) **inside datasets** and b) in data **file names** for transfer of data/results between analytic systems. All values are unique."),
    locations = gt::cells_column_labels(
      columns = location_code
    )
  ) %>%
  gt::tab_footnote(
    footnote = gt::md("**location_name**: Used (rarely) **inside results** (figures, tables, documents). Can be confusing as some names are duplicated."),
    locations = gt::cells_column_labels(
      columns = location_name
    )
  ) %>%
  gt::tab_style(
    style = list(
      gt::cell_text(decorate = "line-through")
      ),
    locations = gt::cells_body(
      columns = location_name,
      rows = gt::everything()
    )
  ) %>%
  gt::tab_footnote(
    footnote = gt::md("**location_name_description_nb**: Used (frequently) **inside results** (figures, tables, documents). All values are unique."),
    locations = gt::cells_column_labels(
      columns = location_name_description_nb
    )
  ) %>%
  gt::tab_footnote(
    footnote = gt::md("**location_name_file_nb_utf**: Used (frequently) in the **file names** for results (figures, tables, documents). All values are unique."),
    locations = gt::cells_column_labels(
      columns = location_name_file_nb_utf
    )
  ) %>%
  gt::tab_footnote(
    footnote = gt::md("**location_name_file_nb_ascii**: Used (rarely) in the **file names** for results (figures, tables, documents). Used if file systems have problems with the Norwegian letters æøå. All values are unique."),
    locations = gt::cells_column_labels(
      columns = location_name_file_nb_ascii
    )
  ) %>%
    gt::tab_footnote(
    footnote = "Bo- og arbeidsmarkedsregioner. Housing and labor market regions.",
    locations = gt::cells_body(
      columns = granularity_geo,
      rows = granularity_geo == "baregion"
    )
  ) %>%
    gt::tab_footnote(
    footnote = "Mattilsynet-regioner. Food authority regions.",
    locations = gt::cells_body(
      columns = granularity_geo,
      rows = granularity_geo == "faregion"
    )
  ) 
```

## Location (reference)

Locations can be obtained from [spldata](/spldata). Valid locations (and location types) are available in `spldata::norway_locations_names()`.

Here we list as a reference table the valid `location_code`s and `location_name_description_nb`s (the two most commonly used locations).

```{r echo=FALSE, results='asis'}
d <- spldata::norway_locations_names()[,.(
  location_order = paste0("#", location_order),
  location_code,
  location_name_description_nb
)]

gt::gt(d) %>%
  gt::tab_options(
    table.width = "750px"
  ) %>%
  gt::tab_header(title = "Reference table of valid `location_code`s and `location_name_description_nb`s") %>%
  gt::cols_label(
    location_order = "#"
  ) 
```

## Ages

Ages should be coded as characters and should always contain 3 digits. If it is an age range, the two ages are joined by a hyphen (e.g. `005-010`). 

Use `085p` instead of `>=085` or `85+`. 

```{r echo=FALSE, results='asis'}
d <- rbind(
  data.frame(
    value = "\"000\"",
    class = "character",
    definition = "One year age group (0 year olds)"
  ),
  data.frame(
    value = "\"079\"",
    class = "character",
    definition = "One year age group(79 year olds)"
  ),
  data.frame(
    value = "\"000-004\"",
    class = "character",
    definition = "Age span of 0-4 year olds"
  ),
  data.frame(
    value = "\"065p\"",
    class = "character",
    definition = "Age span of >=65 year olds"
  ),
  data.frame(
    value = "NA_character_",
    class = "character",
    definition = "Missing/unknown"
  ),
  data.frame(
    value = "\"total\"",
    class = "character",
    definition = "Everyone"
  )
)

gt::gt(d) %>%
  gt::tab_header(title = "Valid ages in the splverse format") %>%
  gt::cols_label(
    value = "Value",
    definition = "Definition"
  )
```

This format will help your data be easily sorted and kept in the right order.

Missing ages should be coded as `NA_character_`.

## Sex

Sex should be coded as characters.

```{r echo=FALSE, results='asis'}
d <- rbind(
  data.frame(
    value = "\"male\"",
    class = "character",
    definition = "Male"
  ),
  data.frame(
    value = "\"female\"",
    class = "character",
    definition = "Female"
  ),
  data.frame(
    value = "NA_character_",
    class = "character",
    definition = "Missing/unknown"
  ),
  data.frame(
    value = "\"total\"",
    class = "character",
    definition = "Everyone"
  )
)

gt::gt(d) %>%
  gt::tab_header(title = "Valid sexes in the splverse format") %>%
  gt::cols_label(
    value = "Value",
    definition = "Definition"
  )
```

Missing sexes should be coded as `NA_character_`.

## Unified variables

All datasets in the splverse format splfmt_rts_v1_mini will contain these 10 variables.

Time conversion functions can be found in package [spltime](https://docs.sykdomspulsen.no/spltime). 

```{r echo=FALSE, results='asis'}
d <- rbind(
  data.frame(
    variable = "granularity_time",
    accepted_values = gt::html("<a href='#Time'>Time</a> (day, isoweek, isoyear)"),
    definition = "Granularity of time"
  ),
  data.frame(
    variable = "granularity_geo",
    accepted_values = paste0(unique(spldata::norway_locations_names()$granularity_geo), collapse=", "),
    definition = "Granularity of geography"
  ),
  data.frame(
    variable = "country_iso3",
    accepted_values = "nor, den, swe, fin",
    definition = "ISO3 country code, e.g. nor for Norway"
  ),
  data.frame(
    variable = "location_code",
    accepted_values = gt::html("<a href='#Location'>Location</a> (norge, countyXX, municipXXXX, ...)"),
    definition = "ISO3 country code, e.g. nor for Norway"
  ),
  data.frame(
    variable = "border",
    accepted_values = "2020",
    definition = "The borders (kommunesammenslåing) that location_code represents"
  ),
  data.frame(
    variable = "age",
    accepted_values = gt::html("<a href='#Age'>Age</a> (0, 1, 000-004, 065p, total, ...)"),
    definition = "Age in years"
  ),
  data.frame(
    variable = "sex",
    accepted_values = gt::html("<a href='#Sex'>Sex</a> (male, female, total)"),
    definition = "Sex"
  ),
  data.frame(
    variable = "isoyear",
    accepted_values = "YYYY",
    definition = "Use function spltime::isoyear_n"
  ),
  data.frame(
    variable = "isoyearweek",
    accepted_values = "\"YYYY-WW\"",
    definition = "Use function spltime::isoyearweek_c"
  ),
  data.frame(
    variable = "date",
    accepted_values = "YYYY-MM-DD",
    definition = "Always corresponds to the last date in the time period. E.g. if granularity_time=='isoweek' then date is the Sunday of that week. If granularity_time == 'event_*_date1_to_9999_01_01' then date is 9999-01-01"
  )
) %>%
  dplyr::mutate(
    accepted_values = purrr::map(accepted_values, ~ gt::html(as.character(.)))
  )

gt::gt(d) %>%
  gt::tab_options(
    table.width = "750px"
  ) %>%
  gt::cols_width(
    variable ~ "20%",
    accepted_values ~ "40%",
    definition ~ "40%"
  ) %>%
  gt::cols_align(
    align = "left"
  ) %>%
  gt::tab_header(title = "10 Unified variables in the splverse format splfmt_rts_v1_mini") %>%
  gt::cols_label(
    variable = "Variable",
    accepted_values = "Accepted values",
    definition = "Definition"
  )
```

## Task-specific variables 

Variable names that contain values should be in one the following forms:

- `descriptive_format`
- `descriptive_format_functiontime`

```{r echo=FALSE, results='asis'}
d <- rbind(
  data.frame(
    x = "",
    examples = "deaths, consultations, cases",
    definition = "Simple"
  ),
  data.frame(
    x = "",
    examples = "hospital_deaths, sysvak_vax_dose_1",
    definition = "Complex"
  ),
  data.frame(
    x = "",
    examples = "msis_cases_testdate, msis_cases_regdate",
    definition = "Complex"
  ),
  
  data.frame(
    x = "",
    examples = "tag",
    definition = "Used when data is in long format, to indicate an id variable."
  ),
  data.frame(
    x = "",
    examples = "n",
    definition = "Integer count"
  ),
  data.frame(
    x = "",
    examples = "cum_n",
    definition = "Integer count corresponding to all time prior"
  ),
  data.frame(
    x = "",
    examples = "pr1",
    definition = "Proportion (between 0 and 1)"
  ),
  data.frame(
    x = "",
    examples = "pr100",
    definition = "Percentage (between 0 and 100)"
  ),
  data.frame(
    x = "",
    examples = "pr100000, prX",
    definition = "Rate per X"
  )
)

gt::gt(d) %>%
  gt::tab_options(
    table.width = "750px"
  ) %>%
  gt::cols_width(
    x ~ "10%",
    examples ~ "30%",
    definition ~ "60%"
  ) %>%
  gt::cols_align(
    align = "left"
  ) %>%
  gt::tab_header(title = "Task-specific variables in the splverse format splfmt_rts_v1_mini") %>%
  gt::cols_label(
    x = "",
    examples = "Examples",
    definition = "Definition"
  ) %>%
  gt::tab_row_group(
    label = "Format",
    rows = 4:nrow(d)
  ) %>%
  gt::tab_row_group(
    label = "Descriptive",
    rows = 1:3
  )
```

#### Tag descriptive

When creating a database table with many different sources of data, it is recommended to include the data source as a descriptive:

- `msis_cases_testdate_n` - number of cases by testing date
- `msis_cases_regdate_n` - number of cases by registration date
- `msislab_tests_n` - number of tests performed
- `nipar_hospital_main_cause_n` - number of people who were admitted to the hospital with covid-19 as a main cause
- `nipar_icu_n` - number of people who were admitted to the ICU with covid-19
- `mor_deaths_n` - number of deaths

#### Tag format

`format` is the format that the value is stored in.

For **percentages/proportions**, we recommend using `pr100` (i.e. values going between 0 and 100) over `pr1` (i.e. values going between 0 and 1).

For **intervals**, use `XintervalY`. We recommend:

- `X` describing the type of interval. If you are using frequentist statistics, this is probably `predinterval` or `confinterval`. If you are using Bayesian statistics, then `credinterval` would be appropriate, and the `descriptive` would then determine if the credible interval is describing a variable (Bayesian prediction interval) or a population/distribution parameter (Bayesian confidence interval).
- `Y` describes the limit, for example (in percentiles) `_p02x5`, `_p97x5`, `_p00x1`, `_p99x9` or (in z-scores) `_l2`, `_u2`, `_l4`, `_u4`.

Some examples are as follows:

- `deaths_n` - number of deaths
- `deaths_n_sum0_13` - number of deaths that occurred today and the preceeding 13 days
- `deaths_pr100000_sum0_13` - number of deaths that occurred today and the preceeding 13 days, per 100 00 population
- `consultations_r80_vk_ote_n` - number of influenza consultations
- `deaths_registered_n` - number of deaths registered
- `deaths_corrected_n` - number of deaths (corrected for registration delay) 
- `deaths_corrected_n_credinterval_p02x5` - Bayesian prediction interval for the "true" number of deaths 
- `deaths_corrected_n_credinterval_p97x5` - Bayesian prediction interval for the "true" number of deaths 
- `deaths_corrected_baseline_n` - Expected number of deaths (i.e. the baseline)
- `deaths_corrected_baseline_n_credinterval_p02x5` - Bayesian prediction interval for the expected number of death (baseline)
- `deaths_corrected_baseline_n_credinterval_p97x5` - Bayesian prediction interval for the expected number of death (baseline)

#### Tag functiontime

We can also include variables that refer to multiple days/weeks:

- `msis_cases_testdate_n_sum0_13` - the aggregate number of cases for today and the preceeding 13 days by testing date
- `msis_cases_testdate_pr100000_sum0_13` - the aggregate number of cases for today and the preceeding 13 days by testing date per 100 000 population
